D3064R0: How to Avoid OOTA Without Really Trying

### 2024-02-24 cuttingroom.tex
February 23 feedback from Alan Stern
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

>>>>

@@@  disagreements.

\begin{figure}[tb]
\begin{center}
\resizebox{4in}{!}{\includegraphics{coe-nvals}}
\caption{On x86, Reasonable CPUs Can Disagree}
\label{fig:On x86; Reasonable CPUs Can Disagree}
\end{center}
\end{figure}

Figure~\ref{fig:On x86; Reasonable CPUs Can Disagree}
shows the results of 79 of the systems's hardware threads simultaneously
storing their identifying integer (ranging from 0 to 78) to a shared
variable, then repeatedly doing timestamped reads from that variable.
The x-axis displays time in TSC periods, each of which is about 0.5~ns
in duration.
The y-axis shows the number of distinct opinions that the hardware
threads have as a function of time, a number that is frequently rather
larger than one.
The backwards-in-time coe link from Thread~1 to Thread~0 in
Figure~\ref{fig:IPC Diagram for coe, fre, and rfe}
is therefore entirely plausible.

\begin{figure}[tb]
\begin{center}
\resizebox{3in}{!}{\rotatebox{90}{\includegraphics{store15tred}}}
\caption{On Power5, coe Links Are Also Partially Ordered}
\label{fig:On Power5, coe Links Are Also Partially Ordered}
\end{center}
\end{figure}

### 2024-02-24 ootavacuous.tex
February 23 feedback from Alan Stern
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

Finally, there might be a desire for hard evidence that coe and fre
links really can go backwards in time.
We provide this evidence on x86 to demonstrate that these effects
are not confined to weakly ordered systems.
This system is a dual-socket system with
Intel(R) Xeon(R) Gold 6138 CPU @ 2.00GHz,
each socket with 20~cores, each having a pair of hardware
threads, for a grand total of 80 hardware threads.
The code generating this data may be found in the \path{CodeSamples/cpu}
directory of the
perfbook~\cite{McKenney2018ParallelProgramming-2018-12-08a}.\footnote{
	\co{git clone git://git.kernel.org/pub/scm/linux/kernel/git/paulmck/perfbook.git}}

\begin{figure}[tb]
\begin{center}
\resizebox{4in}{!}{\includegraphics{coe-nvals}}
\caption{On x86, Reasonable CPUs Can Disagree}
\label{fig:On x86; Reasonable CPUs Can Disagree}
\end{center}
\end{figure}

Figure~\ref{fig:On x86; Reasonable CPUs Can Disagree}
shows the results of 79 of the systems's hardware threads simultaneously
storing their identifying integer (ranging from 0 to 78) to a shared
variable, then repeatedly doing timestamped reads from that variable.
The x-axis displays time in TSC periods, each of which is about 0.5~ns
in duration.
The y-axis shows the number of distinct opinions that the hardware
threads have as a function of time, a number that is frequently rather
larger than one.
The backwards-in-time coe link from Thread~1 to Thread~0 in
Figure~\ref{fig:IPC Diagram for coe, fre, and rfe}
is therefore entirely plausible.

@@

Not only is this plausible, but as you can see in
Figure~\ref{fig:On x86; coe Links Are Atemporal},
it actually happens, and frequently.
The store-to-store latency is measured from the end of the ``winning''
store to the beginning of the nonwinning store that started latest.
Every data point on the negative x-axis represents a group of runs
in which the winning store was not the last store.
Thus, the winning store not being the last store is quite common, even
on x86.

\begin{figure}[tb]
\begin{center}
\resizebox{5in}{!}{\includegraphics{coe-out-1}}
\caption{On x86, coe Links Are Partially Ordered}
\label{fig:On x86, coe Links Are Partially Ordered}
\end{center}
\end{figure}

\begin{figure}[tb]
\begin{center}
\resizebox{3in}{!}{\rotatebox{90}{\includegraphics{store15tred}}}
\caption{On Power5, coe Links Are Also Partially Ordered}
\label{fig:On Power5, coe Links Are Also Partially Ordered}
\end{center}
\end{figure}

However, Figure~\ref{fig:On x86, coe Links Are Partially Ordered}
shows that the order of reads of each value at each thread are
consistent with a number of global orders, as required.\footnote{
	Taken on a 16-CPU x86 laptop to avoid an unreadable
	diagram containing 79 bubbles.}
Figure~\ref{fig:On Power5, coe Links Are Also Partially Ordered}
shows a rather more elaborate partial order from a Power5
system~\cite{McKenney2018ParallelProgramming-2018-12-08a}.

@@

Figure~\ref{fig:Propagation Delay and Temporal Properties of coe, fre, and rfe}
shows how the temporal properties of coe, fre, and rfe are explained
by propagation delay.
In the case of coe, the two stores take time to propagate across the
system, and their ordering can be and often is decided after the fact.
In the case of fre, the store again takes time to propagate across
the system, and so a load might execute later in global time than
did the store, but not late enough to return that store's value.
In the case of rfe, the load cannot return a given store's value until
the value has propagated to that part of the system executing that load.

>>>>

Finally, there might be a desire for hard evidence that coe and fre
links really can go backwards in time.
We provide this evidence on x86 to demonstrate that these effects
are not confined to weakly ordered systems.
This system is a dual-socket system with
Intel(R) Xeon(R) Gold 6138 CPU @ 2.00GHz,
each socket with 20~cores, each having a pair of hardware
threads, for a grand total of 80 hardware threads.
The code generating this data may be found in the \path{CodeSamples/cpu}
directory of the git archive of the book
``Is Parallel Programming Hard, And, If So, What Can You Do About
It?''~\cite{McKenney2018ParallelProgramming-2018-12-08a}.\footnote{
	\co{git://git.kernel.org/pub/scm/linux/kernel/git/paulmck/perfbook.git}}

@@

Figure~\ref{fig:On x86; coe Links Are Atemporal}
shows that coe links really are atemporal, that is, they can go backwards
in time, and quite frequently at that, even on x86.
The store-to-store latency is measured from the end of the ``winning''
store to the beginning of the nonwinning store that started latest.
Every data point on the negative x-axis represents a group of runs
in which the winning store was not the last store.
Thus, the winning store not being the last store is quite common, even
on x86.

\begin{figure}[tb]
\begin{center}
\resizebox{5in}{!}{\includegraphics{coe-out-1}}
\caption{On x86, coe Links Form a Coherent Partial Order}
\label{fig:On x86, coe Links Form a Coherent Partial Order}
\end{center}
\end{figure}

However, Figure~\ref{fig:On x86, coe Links Form a Coherent Partial Order}
shows that the order of reads of each value at each thread are
consistent with a number of global orders, as required.\footnote{
	Note that this data was not measured on the 80-CPU system, but
	rather on a 16-CPU x86 laptop to avoid an unreadable diagram
	containing 79 bubbles.}
Working from the far left, one CPU saw the values 0, 11, and 9,
a second CPU saw 3, 11, and 9, and so on.
Although different CPUs saw different sequences of values, there is no
disagreement on the order of the values that they could see.
For example, all CPUs agree that the value 9 came last.
This is an example of single-variable sequential consistency, and
meets the coherence requirements of the C++ standard.

@@

Figure~\ref{fig:Propagation Delay and Temporal Properties of coe, fre, and rfe}
shows how the temporal properties of coe, fre, and rfe are explained
by propagation delay.
The upper portion of the figure shows an atemporal coe, in which
CPU~0 stores 1 to \co{x} after CPU~3 stores 2 to \co{x}, but
the ordering of the stores is decided after the fact.
In this case, the value of 2 from CPU~3's store overwrites the value
of 1 from CPU~0's store, despite the fact that CPU~0's store
happened later in time.
The middle portion of the figure shows an atemporal fre, in which
CPU~3's load obtains the old value of zero from \co{x} despite
having executed long after CPU~0's store, courtesy of the fact
that the new value of \co{x} had not yet propagated to CPU~3.
The lower portion of the figure shows a temporal rfe.
Because CPU~3's load cannot obtain the value stored by CPU~0 until
after that value has propagated to CPU~3, the fact that CPU~3 obtains
the new value implies that the load was executed after the store.
